\documentclass[12pt]{article}


\usepackage{amsmath,amssymb,amsfonts}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{geometry}
\usepackage[numbers,super,sort&compress]{natbib}
\usepackage{booktabs}
\usepackage{multirow}
\geometry{margin=1in}


% Hyperlinks (invisible style: normal text color; no boxes)
\usepackage[hidelinks]{hyperref}
\hypersetup{linktoc=all}
\usepackage[capitalize,noabbrev,nameinlink]{cleveref}


\newtheorem{definition}{Definition}[section]
\newtheorem{assumption}[definition]{Assumption}
\newtheorem{remark}[definition]{Remark}
\newtheorem{proposition}[definition]{Proposition}
\newtheorem{example}[definition]{Example}




\begin{document}


% ==================================================================
% STYLE GUIDELINES
% ==================================================================
% 1. State object and claim once; no meta-commentary or repetition.
% 2. Lead with operational target (what to compute/test), then formalism.
% 3. Humility is informational: scope conditions and failure modes, not hedging.
% 4. Compact orientation aids (one toy example; short roadmap), then move on.
% 5. Empirical tests specify: what we compute, expected result, implications.
% 6. Timelessness: no version numbers in prose; frame progress as cumulative.
% 7. Nulls as constraints: non-findings are model-space restrictions, not failures.
% ==================================================================


\begin{center}
\textit{This document is a technical specification for an ongoing research program.}
\end{center}


\section{Introduction}
\label{sec:introduction}


Public discussion of automation and AI is surrounded by anxiety about job loss and wage stagnation. Important empirical questions include how to measure which parts of the labor market are exposed to technological change, how exposure propagates beyond an initial cluster of tasks, and which measurement choices survive falsifiable checks. Existing task-based exposure measures---routine versus non-routine, ``AI-exposed'' versus not---provide useful summaries but do not directly encode graded similarity across tasks, institutional barriers to worker reallocation, or spillovers from a shock that originates in one region of the task space and affects nearby tasks.


This paper develops a geometric framework for measuring occupational mobility and applies it to study how workers move across the labor market. The central theoretical contribution is the decomposition of effective distance between occupations into semantic and institutional components (\cref{subsec:decomposition}). Semantic distance captures task-content similarity: can the worker perform the tasks? Institutional distance captures credentialing and training barriers: is the worker allowed to perform the job? Two occupations may be semantically similar (e.g., brain surgeon and butcher both involve cutting) but economically distant due to vast differences in required credentials.


Our research program has five contributions thus far. First, we demonstrate that task space has metric structure relevant for labor market outcomes. Wasserstein distance---measuring the minimum cost to transform one occupation's skill portfolio into another---substantially outperforms kernel overlap for predicting worker mobility ($\Delta \text{log-likelihood} = +9{,}576$). This validates the ``earth mover's'' interpretation of occupational distance: workers minimize skill transformation cost when switching occupations.

Second, we validate the semantic-institutional decomposition using worker-level occupation transitions from the Current Population Survey. Both components are significant predictors of observed mobility. Notably, the choice of semantic measure affects the decomposition: Wasserstein distance correlates more strongly with institutional distance ($r = 0.25$) than kernel overlap does ($r = 0.09$), indicating that Wasserstein absorbs variance from the institutional channel and yields a cleaner decomposition where the residual institutional component captures non-skill barriers (credentialing gates, licensing, discrimination).

Third, we test whether institutional barriers are asymmetric---whether credentials function as ``one-way gates'' that restrict entry but not exit. Simple credentialing theory predicts upward barriers should dominate: a surgeon can become a butcher, but not the reverse. We find the opposite: upward and downward barriers are nearly symmetric ($\beta_{\mathrm{up}}/\beta_{\mathrm{down}} = 1.04$). This is an informative constraint on model space: the symmetric barrier finding rules out pure ``one-way gate'' theories of credentialing. O*NET Job Zones capture general occupational distance rather than regulatory gatekeeping; overqualification discrimination and human capital specificity create genuine downward barriers \citep{Pedulla2016,GathmannSchonberg2010}.

Fourth, we apply the framework to automation prediction. Semantic exposure adds $\Delta R^2 = 2.2\%$ over canonical routine task intensity (RTI) in predicting 2019--2024 employment changes ($p = 0.075$). The framework succeeds at measuring task similarity but shows only marginal improvement over existing measures for predicting automation-driven outcomes.

Fifth, we document methodological infrastructure: kernel bandwidth must be calibrated to local distance structure, and RTI must use the full 16-element Acemoglu-Autor composite---single O*NET elements yield $R^2 \approx 0$.


The primary mobility validation uses 89,329 verified occupation transitions from the CPS. In a conditional logit model of destination choice, Wasserstein-based semantic distance ($\alpha = 8.936$, $t = 206.5$) and institutional distance ($\beta = 0.142$, $t = 34.6$) are both significant predictors of observed destinations. Wasserstein substantially outperforms kernel overlap ($\alpha = 5.688$, $t = 160.3$; $\Delta \text{log-likelihood} = +9{,}576$). A complementary validation using wage comovement shows that semantic overlap predicts occupation-pair wage correlations ($R^2 = 0.00523$), outperforming binary Jaccard overlap by a factor of three.


The framework sits within a broader research program targeting a \emph{task-space oracle}---a hypothetical complete structural mapping from task representations, technology shocks, and institutions to equilibrium labor-market outcomes. We do not estimate or approximate this oracle; we test candidate components against observed data. The Wasserstein validation, symmetric barriers finding, and marginal automation result constrain future specifications: metric structure matters for mobility prediction, effective distance does not require directional asymmetry, and semantic geometry alone is insufficient for automation prediction.


The paper explores a model space defined by three axes: (i) \emph{representation}---how to encode tasks (MPNet embeddings, JobBERT, others); (ii) \emph{geometry}---how to measure occupation distance (Wasserstein, kernel overlap); and (iii) \emph{shocks}---how to characterize technology (generic capability, modality-specific). Each empirical test constrains this space. The Wasserstein validation resolves a key methodological question: optimal transport geometry substantially outperforms kernel overlap. The symmetric barrier finding and marginal automation result are informative constraints on model space, not negative findings: they rule out pure ``one-way gate'' theories of credentialing and establish that semantic geometry alone is insufficient for near-term automation prediction.


\cref{sec:literature} reviews related literatures. \cref{sec:theory} develops the theoretical framework, beginning with the task-space oracle as organizing metaphor and including the semantic-institutional decomposition with both symmetric and asymmetric specifications. \cref{sec:data} operationalizes each component using O*NET data. \cref{sec:validation} reports the CPS mobility validation, the geometry comparison, the asymmetric barriers test, wage comovement, and the automation prediction application. \cref{sec:extensions} records extensions. \cref{sec:conclusion} concludes.


\section{Literature Review}
\label{sec:literature}


Research on automation and labor markets has increasingly adopted a task-based framework, analyzing how technologies complement or displace specific activities within occupations. These approaches can be understood as early attempts at low-dimensional approximations to a complete task-space oracle---coarse models that partition the task domain and predict differential effects across partitions.


In the routine-biased technological change (RBTC) paradigm, computerization disproportionately substitutes for routine, rule-based tasks while complementing non-routine analytic and interactive tasks \citep{AutorLevyMurnane2003}. Subsequent work formalized task assignment and emphasized that technological change acts on tasks, with equilibrium implications for employment reallocation and wages that cannot be recovered from a two-skill canonical model alone \citep{AcemogluAutor2011}. Empirically, this perspective underlies prominent exposure designs in which local labor markets more specialized in routine task work experienced larger subsequent declines in routine employment and a reallocation toward service work \citep{AutorDorn2013}, with cross-country evidence documenting related polarization patterns \citep{GoosManning2007}.


The canonical RBTC empirical methodology, established by \citet{AutorLevyMurnane2003} and extended by \citet{AutorDorn2013}, constructs task measures from the Dictionary of Occupational Titles (DOT) 1977 edition using five specific variables: GED-MATH (quantitative reasoning), DCP (direction/control/planning), STS (set limits/tolerances/standards), FINGDEX (finger dexterity), and EYEHAND (eye-hand coordination). These are aggregated into routine, abstract, and manual task indices, with the routine task intensity (RTI) index defined as the log-ratio $\text{RTI}_k = \ln(R_k) - \ln(M_k) - \ln(A_k)$. This methodology provides the benchmark against which continuous semantic measures can be tested. The key question is whether kernel-weighted task overlap with a routine shock profile predicts canonical outcomes better than discrete RTI-based classification.


The limitation, for measurement, is that many task-based implementations operationalize exposure through discrete partitions. Those partitions capture first-order patterns, but they obscure two objects that matter for reduced-form exposure regressions. First, within-category heterogeneity can be large: occupations classified as ``routine'' differ in how close their actual task bundles are to the frontier of a given technology. Second, substitution and complementarity are naturally graded: effects propagate most strongly to nearby tasks, not only to tasks that share a label. A potential contribution of this framework is to define a metric on the task/activity domain itself and represent occupations as distributions over that domain, so that proximity and overlap in task space determine exposure propagation.


This concern is not purely conceptual. Work using richer task detail documents economically meaningful variation within occupations and within broad task categories \citep{AutorHandel2013,SpitzOener2006}. Related evidence in the automation-risk literature shows that occupation-level exposure measures can substantially overstate the share of employment at high risk when they ignore within-occupation task heterogeneity \citep{ArntzGregoryZierahn2016}.


A separate literature studies worker mobility and occupational transitions. Structural labor and search models \citep[in the tradition of][]{Dorn2009} treat occupation-to-occupation transitions as revealing information about frictions and substitutability. If workers can easily move between two occupations, those occupations are economically close regardless of their linguistic descriptions. This motivates using observed mobility as a validation target for task-space geometry: a good representation should predict where workers actually go when they switch.


Historical episodes motivate the same object. Electrification illustrates how a general-purpose technology can shift the demand for task bundles in heterogeneous ways, with task-based evidence in early twentieth-century manufacturing consistent with reweighting across skill-relevant activities \citep{Gray2013}. Classic accounts emphasize the emergence of technology--skill complementarity under electrification through reorganized production, capital deepening, and the rising importance of skilled maintenance and management \citep{GoldinKatz1998,GoldinKatz2008}.


A separate literature studies exposure measurement in shift-share and related designs. The baseline empirical object is composition-predicted exposure: pre-period shares (industries, occupations, or tasks) interacted with common shocks. Work on Bartik instruments clarifies when such designs identify causal effects and how their weighting structure maps local exposure to national shocks \citep{GoldsmithPinkhamSorkinSwift2020}. More recent econometric treatments emphasize conditions under which identification can be obtained from shock variation rather than from the exogeneity of shares, along with practical guidance for inference and diagnostic checks at the shock level \citep{BorusyakHullJaravel2022}. This paper does not propose a new identification strategy; it proposes a different exposure regressor (potentially distance-propagated, task-geometry-based) and tests whether it adds content beyond standard discrete and composition-based baselines under familiar identifying assumptions.


Recent work on generative AI provides transparent, task-level occupation exposure indices that score tasks by model capability and aggregate to the occupation level \citep{EloundouEtAl2024}. Their approach operates at high task granularity, but the exposure signal is categorical (e.g., ordered classes) and aggregated via a fixed linear formula. These can be understood as ad hoc oracle proxies: simple similarity scores between occupation descriptors and technology capabilities. The gap we target is different: we model exposure as a continuous function of proximity in a capability/task space and allow ``nearby'' tasks to receive partial exposure weight under an explicit kernel. Our program asks: what structure should such proxies respect?


A growing literature on the mismatch between linguistic similarity and economic substitutability is relevant to interpreting our findings. Text embeddings capture semantic relatedness---words that appear in similar contexts---but economic substitutability depends on whether workers can actually perform different tasks, whether employers consider credentials interchangeable, and whether institutional factors constrain mobility. The semantic-institutional decomposition directly addresses this gap: semantic distance measures task-content similarity, while institutional distance measures credentialing and training barriers that may operate orthogonally.


Finally, measured technological exposure should be separated from distribution and pass-through. Classic and modern evidence documents that productivity growth and compensation growth can diverge for measurement and institutional reasons \citep{BosworthPerry1994,FleckGlaserSprague2011}. More recent work argues that institutional changes affecting bargaining and rents play a first-order role in wage dynamics \citep{StansburySummers2020}. In our framework, exposure coefficients are interpreted as reduced-form effects conditional on pass-through.


Methodologically, the framework sits alongside propagation approaches that model how shocks diffuse across related units using distance-weighted or network-based dependence structures \citep{Conley1999}. The construction of task geometry is also related to text-based matching approaches that map technologies to occupations using semantic similarity \citep{Webb2020}. A key methodological distinction: Webb matches patent text to job descriptions (cross-domain matching), whereas the present paper matches task text to task text within the O*NET structure (intra-domain matching). This distinction matters for the ``Admissible task representation'' requirement (\cref{assump:admissible-representation}): intra-domain matching avoids the challenge of aligning heterogeneous text corpora.




\section{Theory}
\label{sec:theory}


This section defines the theoretical objects, building from primitives to the structural mapping: (i) the task-space oracle as organizing metaphor; (ii) the task domain as a metric-measure space; (iii) occupations as probability measures over tasks; (iv) structural grounding via switching cost models; (v) the semantic-institutional distance decomposition; (vi) optimal transport as primary geometry; and (vii) propagation mechanisms. Throughout, we develop toy models to build intuition before the empirical implementation in \cref{sec:data}.


\subsection{The Task-Space Oracle}
\label{subsec:oracle}


The \emph{task-space oracle} $\mathcal{F}: (\mathcal{T}, \mathcal{S}, \mathcal{I}, \mathcal{M}) \to (\Delta \mathbf{T}, \mathbf{L}, \mathbf{W}, \mathbf{U})$ is the research program target: a complete structural mapping from task-space representation ($\mathcal{T}$), technology shocks ($\mathcal{S}$), institutions ($\mathcal{I}$), and equilibrium mechanisms ($\mathcal{M}$) to task structure changes, employment, wages, and unemployment. 

This paper validates candidate representations for two components:
\begin{itemize}
    \item $\mathcal{T}$ (semantic geometry): Text-embedding-based task similarity using O*NET Detailed Work Activities, with Wasserstein distance as the primary measure
    \item $\mathcal{I}$ (institutional distance): Job Zone and certification barriers as mobility frictions
\end{itemize}
Future work addresses $\mathcal{S}$ (shock profiles---generic capability vs.\ modality-specific) and $\mathcal{M}$ (equilibrium mechanisms---search, matching, learning). The oracle is an organizing metaphor, not an estimation target. Its value is taxonomic: it clarifies which components a complete model would require, allowing us to test components in isolation and identify which combinations survive falsifiable checks.


\subsection{Structural Grounding}
\label{subsec:structural-grounding}


Task-based occupational distance measures are production function specifications, not objects derived from worker optimization problems. Existing measures---including ours---are empirical constructs validated against outcomes. The decomposition $d_{\mathrm{eff}} = d_{\mathrm{sem}} + \lambda \cdot d_{\mathrm{inst}}$ finds structural support in \citet{Traiberman2019}, who parameterizes switching costs as linear in task-space distance with dimension-specific prices.


Traiberman's specification treats switching costs as:
\[
C(o, o') = \exp\left[\Gamma^M + \Gamma^S \cdot \delta_{\mathrm{in\text{-}sector}} + \Gamma^O \cdot \delta_{\mathrm{out\text{-}sector}} + \sum_i (\upsilon^{o'}_i - \upsilon^o_i) \cdot \Gamma_i\right],
\]
where $\upsilon^o$ is occupation $o$'s task intensity vector and $\Gamma_i$ prices movement along each task dimension. This directly accommodates our semantic-institutional decomposition by adding credential indicators as additional cost dimensions.


Identification uses conditional choice probability methods \citep{ArcidiaconoMiller2011}, exploiting that workers moving to identical occupations face identical continuation values. Estimated switching costs average five years of income (interquartile range: two years), with intrasectoral occupation switching 28\% costlier than pure sectoral switching.


\begin{remark}[Relation to other structural models]
\label{rem:structural-comparison}
Earlier structural models provide useful contrasts. \citet{ArtucChaudhuriMcLaren2010} parameterize sector-switching costs at 6.5 times annual wages but treat all transitions equivalently. \citet{KambourovManovskii2009} demonstrate that occupational tenure returns dominate firm and industry returns (12--20\% for five years), establishing that human capital is occupation-specific---but their model assumes binary specificity rather than continuous task distance. \citet{Lazear2009} provides the skill-weights approach: all skills are general but occupations weight them differently, with specificity arising from the combination.
\end{remark}


\subsection{The Task Manifold}
\label{subsec:task-manifold}


We measure task similarity using text embeddings of O*NET work activity descriptions. Activities described in similar language are treated as economically close; activities with dissimilar descriptions are distant. This similarity structure is formalized as a metric space.


\begin{definition}[Task domain]
\label{def:task-domain}
The \emph{task domain} is a compact metric-measure space
\[
\mathcal{T} = (T,d,\mu),
\]
where $T$ is a set of tasks, $d : T \times T \to \mathbb{R}_{\ge 0}$ is a metric encoding similarity, and $\mu$ is a reference measure for economic mass.
\end{definition}


Distance is the primitive notion of similarity: tasks $x,y$ are close if capabilities and complements transfer at low friction. The framework uses this geometry to encode a prior on spillovers: shocks affect nearby tasks more strongly.


\begin{remark}[Manifold vs Graph]
\label{rem:graph-vs-manifold}
The manifold representation assumes continuous similarity: tasks $x, y$ at small distance $d(x,y)$ are economically interchangeable at low friction. An alternative representation treats activities as discrete nodes in a graph, where adjacency (shared occupation usage) rather than continuous distance determines relatedness. The empirical program tests which representation better predicts economic outcomes.
\end{remark}


\begin{example}[The Task Line]
\label{ex:task-line}
Consider a simple task domain $T = [0, 1]$ representing a single dimension of complexity (e.g., ``manual'' to ``cognitive''). The metric $d(x, y) = |x - y|$ captures the cost of retooling from task $x$ to task $y$. The reference measure $\mu$ is uniform, representing an equal density of economic value across the complexity spectrum.
\end{example}


\begin{assumption}[Admissible task representation]
\label{assump:admissible-representation}
An empirical representation of the task domain is admissible if, for at least one validation target not used in construction (e.g., wage comovement, worker mobility), occupation-pair relatedness in activity space predicts stronger observed relatedness. Specifically:
\begin{enumerate}
    \item \textbf{Metric axioms (if continuous).} The distance $\widehat d$ satisfies the metric axioms on its domain.
    \item \textbf{Scale robustness.} The representation is invariant to irrelevant rescalings of inputs.
    \item \textbf{External diagnostic.} For at least one validation target not used in construction, greater activity-space relatedness predicts stronger observed economic relatedness.
\end{enumerate}
\end{assumption}


\subsection{Occupations as Measures}
\label{subsec:occupations}


Each occupation is represented by a probability distribution over activities, constructed from O*NET importance ratings. The distribution encodes how intensively the occupation uses each activity.


\begin{definition}[Occupations as measures]
\label{def:occupations}
An occupation $j$ is represented by a probability measure $\rho_j$ over $T$ with density $f_j(x) = d\rho_j/d\mu(x)$, interpreted as the intensity with which occupation $j$ uses task $x$.
\end{definition}


\begin{example}[Occupations on the Line (toy illustration)]
\label{ex:occupations-line}
To build intuition, consider two occupations on the task line $T = [0,1]$. ``Occupation A'' (Specialist) is concentrated at the manual end, modeled as a Gaussian measure centered at $x=0.2$ with narrow variance $\sigma_A^2 = 0.01$. ``Occupation B'' (Generalist) is centered at $x=0.8$ but has wider variance $\sigma_B^2 = 0.04$, so its support extends partially into the middle range. The overlap integral $\int \rho_A(x) \rho_B(x)\, d\mu(x)$ quantifies their task-space relatedness. This is intuition only; the empirical implementation uses 2,087 activities and 894 occupations.
\end{example}


\subsection{The Semantic-Institutional Decomposition}
\label{subsec:decomposition}


A central theoretical contribution is the decomposition of effective distance between occupations into semantic and institutional components. This addresses the problem that semantic similarity (task-content overlap) does not imply economic proximity (ease of worker reallocation).


\begin{definition}[Effective distance decomposition]
\label{def:effective-distance}
Define the \emph{effective distance} between occupations $i$ and $j$ as:
\[
d_{\mathrm{eff}}(i,j) = d_{\mathrm{sem}}(i,j) + \lambda \cdot d_{\mathrm{inst}}(i,j),
\]
where:
\begin{itemize}
    \item $d_{\mathrm{sem}}(i,j) \geq 0$ is \emph{semantic distance}: the Wasserstein distance between activity distributions (primary) or one minus normalized kernel-weighted overlap (comparison baseline). It measures task-content dissimilarity.
    \item $d_{\mathrm{inst}}(i,j) \geq 0$ is \emph{institutional distance}: the sum of job zone difference and certification barrier difference. It measures credentialing and training dissimilarity.
    \item $\lambda \geq 0$ is the relative weight of institutional vs.\ semantic friction.
\end{itemize}
\end{definition}


\begin{remark}[Semantic operationalization affects decomposition]
\label{rem:operationalization-decomposition}
The choice of semantic distance measure affects the correlation structure with institutional distance:
\begin{itemize}
    \item Kernel overlap: $\rho(d_{\mathrm{kernel}}, d_{\mathrm{inst}}) = 0.088$ (nearly orthogonal)
    \item Wasserstein: $\rho(d_{\mathrm{wasserstein}}, d_{\mathrm{inst}}) = 0.249$ (moderate correlation)
\end{itemize}
This difference has substantive interpretation. Kernel overlap asks: ``Do these occupations share activities?'' This is nearly orthogonal to training requirements---high-training occupations can have zero overlap with each other. Wasserstein asks: ``How costly is the skill transformation?'' This naturally correlates with training requirements, because transforming a low-skill portfolio into a high-skill portfolio requires investment.

The implication: Wasserstein provides a purer measure of skill-based barriers. The residual institutional component (after controlling for Wasserstein distance) captures non-skill barriers---actual credentialing gates, licensing, discrimination, network effects. The decomposition becomes:
\[
d_{\mathrm{eff}} = d_{\mathrm{transform}} + \lambda \cdot d_{\mathrm{barrier}},
\]
where $d_{\mathrm{transform}}$ is skill transformation cost (Wasserstein) and $d_{\mathrm{barrier}}$ is residual non-skill obstacles (Job Zones after controlling for transformation cost).
\end{remark}


\begin{remark}[Theoretical basis for additive separability]
\label{rem:separability}
Additive separability requires three maintained assumptions:
\begin{enumerate}
    \item Skill acquisition costs depend on task similarity ($d_{\mathrm{sem}}$)
    \item Credential costs are orthogonal to skill requirements ($d_{\mathrm{inst}}$)
    \item No interaction between skill learning and credential acquisition
\end{enumerate}
These assumptions are strong---credentials and skills are often bundled (medical licensing requires both knowledge and certification). We maintain separability as a tractable approximation, following \citet{Traiberman2019}'s treatment of task dimensions and sector indicators as additively separable cost components. Identification requires variation where task similarity and credential requirements vary independently; our CPS validation tests whether this decomposition predicts worker transitions.
\end{remark}


\begin{remark}[The brain surgeon / butcher problem]
\label{rem:surgeon-butcher}
Consider two occupations: brain surgeon and butcher. Both involve cutting tissue with precision tools. On a purely semantic basis (task descriptions involving cutting, tool use, manual dexterity), they may appear similar. But economically, they are vastly distant: a butcher cannot become a brain surgeon without years of medical training and credentialing. The semantic distance $d_{\mathrm{sem}}$ may be moderate, but the institutional distance $d_{\mathrm{inst}}$ is maximal (job zone 5 vs.\ job zone 2; extensive certification vs.\ minimal). The effective distance correctly identifies them as economically distant despite superficial task similarity.
\end{remark}


\begin{remark}[Testable implications]
\label{rem:decomposition-test}
If $\lambda > 0$ and statistically significant, institutional distance predicts mobility independently of semantic distance. The CPS mobility test (\cref{subsec:cps-mobility}) estimates $\lambda$ from observed transition patterns.
\end{remark}


\begin{definition}[Asymmetric institutional distance]
\label{def:asymmetric-distance}
The symmetric $d_{\mathrm{inst}}$ can be decomposed into directional components:
\[
d_{\mathrm{inst}}^{\mathrm{up}}(i,j) = \max(0, \mathrm{Zone}_j - \mathrm{Zone}_i) + \gamma \cdot \max(0, \mathrm{Cert}_j - \mathrm{Cert}_i),
\]
\[
d_{\mathrm{inst}}^{\mathrm{down}}(i,j) = \max(0, \mathrm{Zone}_i - \mathrm{Zone}_j) + \gamma \cdot \max(0, \mathrm{Cert}_i - \mathrm{Cert}_j),
\]
where $\mathrm{Zone} \in \{1,\ldots,5\}$ is the O*NET job zone and $\mathrm{Cert}$ is the normalized certification importance. By construction, $d_{\mathrm{inst}}^{\mathrm{up}}(i,j) + d_{\mathrm{inst}}^{\mathrm{down}}(i,j) = d_{\mathrm{inst}}(i,j)$.
\end{definition}


\begin{remark}[Theoretical predictions for asymmetry]
\label{rem:asymmetry-prediction}
Two competing theoretical models yield different predictions:

\textbf{Credential-gate theory} predicts $\beta_{\mathrm{up}} \gg \beta_{\mathrm{down}}$ because licensing restricts entry but not exit \citep{Jackson2023}. A surgeon can become a butcher; the reverse requires medical school. Under this model, Job Zones proxy licensing barriers that create one-way gates.

\textbf{Human capital specificity theory} predicts $\beta_{\mathrm{up}} \approx \beta_{\mathrm{down}}$ because workers accumulate occupation-specific capital lost in any move. Three mechanisms generate symmetric barriers:
\begin{enumerate}
    \item \textbf{Occupation-specific human capital} \citep{KambourovManovskii2009}: 12--20\% wage premiums for five years of occupation tenure are forfeited in any switch.
    \item \textbf{Overqualification stigma} \citep{Pedulla2016}: Employers penalize overqualified applicants with callback penalties approximately as severe as one year of unemployment, creating genuine downward barriers.
    \item \textbf{Task-specific skill destruction} \citep{GathmannSchonberg2010}: Up to 52\% of wage growth derives from task-specific human capital destroyed in downward moves.
\end{enumerate}

\textbf{Distinction:} ``Distance-like'' frictions are continuous, symmetric, and separation-dependent. ``Gate-like'' frictions are discontinuous entry barriers with zero exit cost. \citet{KleinerXu2020} find licensing reduces \emph{both} entry and exit rates by 24\%---contradicting the one-way gate model.

\cref{subsubsec:asymmetric-test} tests these competing predictions empirically.
\end{remark}


\subsection{Optimal Transport as Primary Geometry}
\label{subsec:optimal-transport}


Kernel-weighted overlap measures shared task mass (smoothed intersection). An alternative approach uses optimal transport (OT) to measure minimum transformation cost between task distributions. The empirical comparison (\cref{subsec:geometry-comparison}) demonstrates that Wasserstein distance substantially outperforms kernel overlap for predicting worker mobility, validating OT as the primary geometry \citep{Galichon2016}.


\begin{definition}[Wasserstein distance between occupations]
\label{def:wasserstein}
The Wasserstein distance between occupations $i$ and $j$ is:
\[
W(\rho_i, \rho_j) = \min_{\gamma \in \Pi(\rho_i, \rho_j)} \sum_{a,b} \gamma(a,b) \cdot d(a,b),
\]
where $\gamma$ is a transport plan in $\Pi(\rho_i, \rho_j)$ (the set of joint distributions with marginals $\rho_i$ and $\rho_j$), and $d(a,b)$ is the ground distance between tasks $a$ and $b$.
\end{definition}


\begin{remark}[Economic interpretation of the transport plan]
\label{rem:transport-interpretation}
The transport plan $\gamma(a,b)$ represents optimal skill transformation---the allocation of retraining investment across task pairs. Entry $\gamma(a,b) \cdot d(a,b)$ is the cost of transforming mass at task $a$ into capability at task $b$. Total Wasserstein distance $W(\rho_i, \rho_j)$ is total transformation cost. This ``earth mover's'' interpretation---the cost to reshape one skill distribution into another---is economically intuitive and empirically validated: workers minimize transformation cost when switching occupations (\cref{subsec:geometry-comparison}).
\end{remark}


\begin{remark}[Kernel overlap vs.\ Wasserstein: theoretical contrast and empirical resolution]
\label{rem:kernel-vs-ot}
\begin{enumerate}
    \item \textbf{Kernel overlap} measures shared task mass with distance-weighted smoothing. Robust to ground metric noise via bandwidth regularization. Does not respect metric structure (triangle inequality).
    \item \textbf{Wasserstein} measures minimum transformation cost. Respects metric structure (proper distance). Sensitive to ground metric misspecification---even 5\% outliers can cause poor transport couplings.
\end{enumerate}

\textbf{Theoretical prediction:} If workers' reallocation costs correspond to skill transformation difficulty, Wasserstein should outperform kernel for mobility prediction. If ground metric is noisy (text embeddings imperfectly capture economic substitutability), kernel's implicit regularization may dominate.

\textbf{Empirical resolution:} Wasserstein substantially outperforms kernel overlap ($\Delta \text{log-likelihood} = +9{,}576$; see \cref{subsec:geometry-comparison}). The metric structure of task space is economically relevant: workers minimize transformation cost, not just shared task overlap.
\end{remark}


\subsection{Kernel Overlap as Comparison Baseline}
\label{subsec:kernel-overlap}


While Wasserstein distance is the primary semantic measure, kernel overlap provides a useful comparison baseline and remains appropriate for some applications (e.g., propagation of shocks across tasks).


\begin{definition}[Kernel-weighted overlap]
\label{def:kernel-overlap}
Define occupation-pair proximity as
\[
\mathrm{Overlap}_K(i,j) = \int_T \int_T \rho_i(x)\, \rho_j(y)\, k(d(x,y); \sigma)\, d\mu(x)\, d\mu(y),
\]
where $k(\cdot; \sigma)$ is a kernel function with bandwidth $\sigma$. Normalized overlap controls for concentration effects:
\[
\mathrm{NormOverlap}_K(i,j) = \frac{\mathrm{Overlap}_K(i,j)}{\sqrt{\mathrm{Overlap}_K(i,i) \cdot \mathrm{Overlap}_K(j,j)}}.
\]
Semantic distance (kernel-based) is $d_{\mathrm{sem}}^{\mathrm{kernel}}(i,j) = 1 - \mathrm{NormOverlap}_K(i,j)$.
\end{definition}


\begin{remark}[Bandwidth calibration]
\label{rem:bandwidth-calibration}
A critical methodological finding is that $\sigma$ must be calibrated to the \emph{local} distance structure (median nearest-neighbor distance), not the global distance distribution (median of all pairwise distances). When distances cluster in a narrow band, calibrating $\sigma$ to the global median causes kernel collapse: all weights become approximately equal, destroying the signal.
\end{remark}


\subsection{Technological Fields and Shocks}
\label{subsec:tech-fields}


Technology shifts labor demand through displacement (substitution) and augmentation (complementarity). We represent both as fields over $T$.


\begin{definition}[Technological state]
\label{def:tech-state}
At time $t$, the technological state is a pair of bounded measurable functions
\[
A_t, C_t : T \to \mathbb{R},
\]
where $A_t(x)\ge 0$ is displacement intensity at task $x$ and $C_t(x)>0$ is a complementarity factor at task $x$.
\end{definition}


\begin{definition}[Technological shock profile]
\label{def:shock-field}
Let $I_t \in L^2(T,\mu)$ denote an exogenous innovation input at time $t$.
The profile $I_t(x)$ represents the intensity with which the frontier is pushed at task $x$ by R\&D, investment, or related innovation activity.
\end{definition}


In applications, $I_t$ is constructed and then potentially propagated through a measured geometry. In the prospective evaluation, multiple candidate $I_t$ are compared; the shock mapping is treated as a testable component of measurement rather than as a structural primitive.


\subsection{Propagation of Shocks Across Tasks}
\label{subsec:propagation}


A technology shock to one activity may spill over to similar activities. We model this as a weighted average: each activity's exposure equals a distance-weighted sum of the shock profile, with nearby activities receiving higher weight. The alternative is no spillover: shocks affect only directly-targeted activities.


\begin{definition}[Propagation mechanisms]
\label{def:propagation-mechanisms}
\textbf{(A) Kernel propagation (continuous).} The spillover operator computes a distance-weighted average:
\[
[\widetilde{\mathcal{K}}_d f](x)
=
\frac{\int_T k\bigl(d(x,y)\bigr)\, f(y)\, d\mu(y)}{\int_T k\bigl(d(x,y)\bigr)\, d\mu(y)},
\]
where $k(\cdot)$ is a kernel function that is positive and decreasing in distance.

\textbf{(B) Identity propagation (discrete).} No spillover: $[\mathcal{I} f](x) = f(x)$. Shocks affect only directly-targeted activities.
\end{definition}


Baseline implementation of kernel propagation uses an exponential kernel with a single spillover scale:
\begin{equation}
\label{eq:exp-kernel}
k\bigl(d(x,y); \sigma\bigr)
=
\exp\!\left(- \frac{d(x,y)}{\sigma}\right),
\qquad
\theta = \sigma > 0.
\end{equation}


\subsection{Aggregation to Occupations and Outcome Equations}
\label{subsec:aggregation}


\begin{definition}[Occupation-level exposure functionals]
\label{def:exposure-functionals}
Given $(A_t,C_t)$ and measures $\{\rho_j\}_{j=1}^J$, define
\[
E^{A}_{j,t} \equiv \int_T A_t(x)\, d\rho_j(x),
\qquad
E^{C}_{j,t} \equiv \int_T \log C_t(x)\, d\rho_j(x).
\]
The pipeline is: (i) shock profile $I_t$ on tasks, (ii) propagation $A_t = \mathcal{P} I_t$ across tasks (continuous or discrete), (iii) aggregation via $\rho_j$ to occupation-level exposure.
\end{definition}


\begin{definition}[Discrete exposure]
\label{def:discrete-exposure}
Under identity propagation ($\mathcal{P} = \mathcal{I}$), occupation-level exposure reduces to:
\[
E^A_j = \sum_{a \in T_n} I(a) \cdot \rho_j(a),
\]
which is a weighted sum of shock intensity over the occupation's activity distribution.
\end{definition}


\begin{definition}[Reduced-form outcome equations (partial equilibrium)]
\label{def:reduced-form}
Let $w_{j,t}$ and $L_{j,t}$ denote wages and employment. We evaluate
\[
\Delta \log w_{j,t} = \beta_w E^{A}_{j,t} + \gamma_w E^{C}_{j,t} + \bm{X}_{j,t}'\delta_w + u_{j,t},
\]
\[
\Delta \log L_{j,t} = \beta_L E^{A}_{j,t} + \gamma_L E^{C}_{j,t} + \bm{X}_{j,t}'\delta_L + v_{j,t},
\]
interpreting coefficients as conditional associations used to compare exposure measures.
\end{definition}


\begin{remark}[Implied covariance structure---diagnostic target]
\label{rem:covariance}
If outcomes load on exposure with common sign and shocks are spatially concentrated in task space, then occupation pairs with greater task-distribution overlap should exhibit stronger outcome comovement.

\textbf{Binary overlap.} Define binary overlap as the Jaccard index:
\[
\mathrm{Overlap}_B(i,j) = \frac{|A_i \cap A_j|}{|A_i \cup A_j|},
\]
where $A_j = \{a : \rho_j(a) > 0\}$ is the activity support of occupation $j$.
\end{remark}


\subsection{Wage-setting wedge and pass-through (scope condition)}
\label{subsec:wedge}


Observed wages need not track marginal products one-for-one. We record this as a scope condition rather than a mechanism to be estimated.


\begin{definition}[Occupation-level wage-setting wedge]
\label{def:wedge}
Let $\mathrm{MPL}_{j,t}$ denote the (unobserved) marginal product of labor for occupation $j$. Define $b_{j,t} > 0$ such that
\[
\log w_{j,t} = \log \mathrm{MPL}_{j,t} + \log b_{j,t}.
\]
\end{definition}


Empirically, proxies for wage-setting environments (e.g., union coverage, licensing prevalence, concentration) enter as controls or descriptive correlates. Interpretation of exposure coefficients is conditional on these environments \citep{StansburySummers2020,AutorSalomons2018}.


\subsection{New tasks and an evolving task domain (extension)}
\label{subsec:newtasks}


The baseline holds $(T,\mu)$ fixed. Over longer horizons, outcomes depend on task creation and reweighting \citep{AcemogluRestrepo2018}. This section records extension paths: expanding $T_t$, time-varying $\mu_t$, and time-varying occupation measures $\rho_{j,t}$.




\section{Data Construction and Measurement}
\label{sec:data}


This section operationalizes the theoretical framework using O*NET data. We describe how to (i) construct a discrete approximation of the task manifold, (ii) build occupation measures from ratings, (iii) compute activity distances, (iv) compute Wasserstein distances between occupations, and (v) specify design choices. Empirical validation follows in \cref{sec:validation}.


\subsection{Model Space Axes}
\label{subsec:model-space}


This paper explores a model space defined by three axes:
\begin{enumerate}
    \item \textbf{Representation axis:} How to represent tasks (MPNet embeddings, JobBERT, others)
    \item \textbf{Geometry axis:} How to measure occupation distance (Wasserstein [primary], kernel overlap [baseline])
    \item \textbf{Shock axis:} How to characterize technology (generic capability, modality-specific)
\end{enumerate}

Each empirical test constrains this space. The geometry comparison (\cref{subsec:geometry-comparison}) resolves a key methodological question: Wasserstein substantially outperforms kernel overlap. Validated combinations advance; rejected combinations are pruned.


\subsection{Discrete Approximation of the Manifold}
\label{subsec:discrete-approx}


We use O*NET Detailed Work Activities (DWAs) as the task unit. O*NET provides 2,087 DWAs linked to 894 occupations via task statements. Each occupation's activity distribution is constructed from importance ratings.


\begin{definition}[Empirical activity domain]
\label{def:empirical-activity-domain}
Let $T_n = \{a_1, \ldots, a_n\}$ denote the set of DWAs ($n = 2{,}087$). The reference measure $\mu$ is uniform. The metric $d$ is constructed from text embeddings (\cref{subsubsec:recipes}).
\end{definition}


\begin{remark}[Constructing occupation measures from O*NET]
\label{rem:rho-construction}
The occupation measure $\rho_j$ is constructed from O*NET ratings via the following steps \citep{onet2024dwa,onet2024scales}:


\medskip
\noindent\textbf{(A) Raw weight construction $w_j(a)$:}


\emph{If $T_n = $ GWA (Work Activities file):}
\begin{itemize}
    \item Retrieve Importance $\mathrm{IM}_{j,a} \in [1,5]$ and Level $\mathrm{LV}_{j,a} \in [0,7]$ for each occupation-activity pair.
    \item Baseline: $w_j(a) = \mathrm{IM}_{j,a} \times \mathrm{LV}_{j,a}$ after rescaling both to $[0,1]$.
\end{itemize}


\emph{If $T_n = $ DWA (Detailed Work Activities):}
\begin{itemize}
    \item DWAs lack direct importance ratings; importance is derived via linked tasks.
    \item Let $\mathrm{Tasks}(j, a) = \{\text{tasks in occupation } j \text{ linked to DWA } a\}$.
    \item Baseline aggregator: $w_j(a) = \max_{t \in \mathrm{Tasks}(j,a)} \mathrm{IM}_{j,t}$ (top importance).
\end{itemize}


\medskip
\noindent\textbf{(B) Suppression and missingness rules:}
\begin{itemize}
    \item Set $w_j(a) = 0$ if O*NET flags ``Recommend Suppress $=$ Y'' for that cell.
    \item Set $w_j(a) = 0$ if ``Not Relevant $=$ Y'' (per O*NET's 75\% rule for Level).
    \item If an occupation has fewer than 10 activities with nonzero weight after filtering, flag the occupation.
\end{itemize}


\medskip
\noindent\textbf{(C) Normalization to probability measure:}
\[
\rho_j(a) = \frac{w_j(a)}{\sum_{a' \in T_n} w_j(a')}.
\]
This yields $\rho_j \in \mathcal{P}(T_n)$.


\medskip
\noindent\textbf{(D) Implementation summary (O*NET 30.0):}

The baseline implementation uses:
\begin{itemize}
    \item Files: Task Ratings.xlsx (task importance by occupation), Tasks to DWAs.xlsx (task-to-DWA linkages), DWA Reference.xlsx (DWA definitions).
    \item Coverage: 894 occupations with valid measures after filtering.
\end{itemize}

\medskip
\noindent\textbf{(E) Distribution statistics:}
\begin{itemize}
    \item Sparsity: 99.1\% of occupation-activity weights are zero.
    \item Median activities per occupation: 19 (of 2,087 possible).
    \item Median pairwise Jaccard overlap: 0.0 (most occupation pairs share no activities).
\end{itemize}
\end{remark}


\begin{remark}[Matrix representation of the spillover operator]
\label{rem:matrix-notation}
In discrete instantiation with uniform $\mu$, the spillover operator becomes a matrix $K \in \mathbb{R}^{n \times n}$:
\[
K_{ab} = k(d(a,b); \sigma) = \exp(-d(a,b)/\sigma).
\]
Note: we do \emph{not} row-normalize $K$ in the baseline implementation.
\end{remark}


\subsubsection{Construction recipes for the activity metric}
\label{subsubsec:recipes}


\begin{remark}[Recipes for constructing $\widehat d$ on activities]
\label{rem:recipes}
Instantiate $T$ as a finite set of activities (GWA or DWA). Occupations are probability measures over activities, not points in $T$.
\begin{enumerate}
    \item \textbf{Recipe Y (text-embedding geometry).} Embed activity titles/descriptions using a sentence encoder (all-mpnet-base-v2, 768 dimensions). Define $\widehat d(a,b)$ as cosine distance in embedding space.
    
    \emph{Validation targets:} mobility, wage comovement.
    
    \item \textbf{Binary Jaccard (no distance).} Define activity-space relatedness via set intersection without any distance weighting:
    \[
    \mathrm{Jaccard}(i,j) = \frac{|A_i \cap A_j|}{|A_i \cup A_j|},
    \]
    where $A_j = \{a : \rho_j(a) > 0\}$.
\end{enumerate}
\end{remark}


\subsection{Wasserstein Distance Computation}
\label{subsec:wasserstein-computation}


Given the activity-level ground metric $d(a,b)$ and occupation measures $\rho_i, \rho_j$, the Wasserstein distance is computed via linear programming. Implementation uses the Python Optimal Transport (POT) library.


\begin{remark}[Computational details]
\label{rem:wasserstein-computation}
\begin{itemize}
    \item \textbf{Sparsity exploitation:} With 99.1\% sparsity, effective transport problems involve only activities with nonzero weight. Median support union: 38 activities per occupation pair.
    \item \textbf{Computation time:} Full $894 \times 894$ matrix computed in 84 seconds.
    \item \textbf{Metric verification:} Triangle inequality verified on random samples; Wasserstein satisfies metric axioms (kernel overlap does not).
\end{itemize}
\end{remark}


\subsection{Design Choices and Pre-Commitment}
\label{subsec:design-choices}


All design choices are fixed before examining outcome data.

\begin{table}[h!]
\centering
\caption{Design Axis Specifications}
\label{tab:design-axes}
\begin{tabular}{lll}
\toprule
\textbf{Axis} & \textbf{Primary} & \textbf{Comparison/Robustness} \\
\midrule
Activity domain & DWA ($n = 2{,}087$) & GWA ($n \approx 41$) \\
Reference measure & Uniform & Hierarchy-weighted \\
Weight construction & Top importance over linked tasks & Mean importance \\
Activity distance & Text embedding (MPNet) & Binary Jaccard \\
\textbf{Occupation distance} & \textbf{Wasserstein} & Kernel overlap \\
Kernel (for propagation) & Exponential, unnormalized & Gaussian \\
Bandwidth & Median NN distance & Grid search \\
\bottomrule
\end{tabular}
\end{table}


\section{Empirical Validation}
\label{sec:validation}


This section reports empirical tests of the semantic-institutional decomposition. Each test follows the structure: objective, specification, pre-committed interpretations, and results. Tests are organized as stress tests of candidate oracle modules.

\textbf{Completed tests:}
\begin{itemize}
    \item Geometry comparison (\cref{subsec:geometry-comparison}): Wasserstein vs.\ kernel overlap
    \item CPS worker mobility (\cref{subsec:cps-mobility}): Tests whether $d_{\mathrm{eff}} = d_{\mathrm{sem}} + \lambda \cdot d_{\mathrm{inst}}$ predicts destination choice
    \item Asymmetric barriers (\cref{subsubsec:asymmetric-test}): Tests whether $\beta_{\mathrm{up}} > \beta_{\mathrm{down}}$
    \item Wage comovement (\cref{subsec:wage-comovement}): Tests whether semantic overlap predicts wage correlations
    \item Automation prediction (\cref{subsec:automation-prediction}): Tests whether semantic exposure improves on RTI
\end{itemize}

\textbf{Pending tests:}
\begin{itemize}
    \item Embedding comparison: JobBERT vs.\ MPNet
    \item Retrospective construct validity against Autor-Dorn RTI
\end{itemize}


\subsection{Primary Validation: Geometry Comparison}
\label{subsec:geometry-comparison}


This section reports the comparison between Wasserstein distance and kernel overlap for predicting worker mobility. The test directly evaluates whether metric structure (optimal transport) outperforms smoothed intersection (kernel overlap).


\subsubsection{Specification}

Both specifications use the same conditional logit framework with 89,329 CPS occupation transitions:
\[
U_{kj} = \alpha \cdot (-d_{\mathrm{sem}}(i,j)) + \beta \cdot (-d_{\mathrm{inst}}(i,j)) + \varepsilon_{kj}
\]

The only difference is the semantic distance measure: Wasserstein vs.\ kernel overlap.


\subsubsection{Results}

\begin{table}[h!]
\centering
\caption{Geometry Comparison: Wasserstein vs.\ Kernel Overlap}
\label{tab:geometry-comparison}
\begin{tabular}{lccccc}
\toprule
\textbf{Specification} & $\alpha$ (semantic) & $t$-stat & $\beta$ (institutional) & $t$-stat & Log-likelihood \\
\midrule
Wasserstein (primary) & 8.936 & 206.5 & 0.142 & 34.6 & $-183{,}051$ \\
Kernel overlap (baseline) & 5.688 & 160.3 & 0.278 & 71.0 & $-192{,}627$ \\
\midrule
Difference & $+57\%$ & --- & $-49\%$ & --- & $+9{,}576$ \\
\bottomrule
\end{tabular}
\end{table}


\subsubsection{Interpretation}

\textbf{Wasserstein substantially outperforms kernel overlap.} The log-likelihood improvement ($\Delta = +9{,}576$) far exceeds pre-committed thresholds (adopt OT if $\Delta > 100$). This validates the ``earth mover's'' interpretation: workers minimize skill transformation cost when switching occupations, not just shared task overlap.

\textbf{The $\beta$ coefficient absorption is theoretically correct.} The institutional coefficient drops from 0.278 to 0.142 ($-49\%$). This reflects the correlation structure: $\rho(d_{\mathrm{wasserstein}}, d_{\mathrm{inst}}) = 0.249$ vs.\ $\rho(d_{\mathrm{kernel}}, d_{\mathrm{inst}}) = 0.088$. Wasserstein absorbs variance that kernel overlap missed---variance related to skill transformation costs that Job Zones partially capture.

The implication: with Wasserstein as the semantic measure, the residual institutional component ($\beta = 0.142$, still significant at $t = 34.6$) captures genuinely non-skill barriers---credentialing gates, licensing requirements, discrimination, network effects. The decomposition becomes cleaner:
\[
d_{\mathrm{eff}} = d_{\mathrm{transform}} + \lambda \cdot d_{\mathrm{barrier}}.
\]

\textbf{Why is the improvement so large?} Kernel distances are compressed near 1.0 (most pairs have approximately zero overlap), while Wasserstein provides genuine variation across the full range. The model has more signal to work with. Additionally, kernel overlap does not respect metric structure (triangle inequality), while Wasserstein does---this matters for occupation choice, where relative distances affect substitution patterns.


\subsection{Worker Mobility and Institutional Wedge}
\label{subsec:cps-mobility}


This section reports the primary empirical validation: a conditional logit model of worker destination choice using CPS occupation transitions. The test directly evaluates whether the semantic-institutional decomposition predicts where workers actually move.


\subsubsection{Data Construction}
\label{subsubsec:cps-data}


\textbf{Source:} IPUMS CPS Basic Monthly Files, 2015-01 through 2019-12 and 2022-01 through 2024-09. The period 2020-01 through 2021-12 is excluded due to COVID labor market disruption. 83 of 96 requested monthly files were obtained.

\textbf{Variables:} CPSIDP (person linking), CPSIDV (validated linking), MISH (month-in-sample), OCC2010 (harmonized occupation code), EMPSTAT (employment status), AGE, SEX, RACE, EDUC (demographics), STATEFIP (geography).


\begin{table}[h!]
\centering
\caption{CPS Sample Construction Pipeline}
\label{tab:cps-pipeline}
\begin{tabular}{lrrp{5cm}}
\toprule
\textbf{Stage} & \textbf{Records} & \textbf{Retention} & \textbf{Notes} \\
\midrule
Raw CPS extract & 10,072,119 & 100\% & 83 monthly samples \\
Employment filter & 4,407,432 & 43.8\% & EMPSTAT $\in$ \{10, 12\}, age 18--65 \\
Consecutive month pairs & 2,353,725 & 53.4\% & CPSIDP linked $t$ to $t+1$ \\
Demographic validation & 2,337,740 & 99.3\% & Age $\pm 1$, sex/race constant \\
Raw transitions identified & 157,606 & 6.74\% & OCC$(t) \neq$ OCC$(t-1)$ \\
Persistence filter & 106,116 & 67.3\% & Verified transitions \\
O*NET mapping successful & 89,329 & 84.2\% & Census 2010 $\to$ O*NET crosswalk \\
\bottomrule
\end{tabular}
\end{table}


\textbf{Measurement error correction.} Raw CPS occupation switching rates ($\sim$6.7\% monthly) substantially exceed true mobility rates ($\sim$0.5--1.0\% per Kambourov \& Manovskii 2008). The discrepancy arises from coding error: self-reported occupation varies across interviews even when the underlying job is unchanged.

We apply a \emph{persistence filter}:
\begin{enumerate}
    \item \textbf{Origin stability:} OCC$(t-2)$ = OCC$(t-1)$ (worker was in origin for at least one prior period)
    \item \textbf{Transition occurred:} OCC$(t) \neq$ OCC$(t-1)$
    \item \textbf{Destination persistence:} OCC$(t)$ = OCC$(t+1)$ = OCC$(t+2)$ (worker remains in destination)
\end{enumerate}
This removes spurious transitions where occupation codes fluctuate due to reporting inconsistency. Retention rate (67.3\%) is consistent with literature estimates that 25--50\% of raw CPS occupation changes are measurement error.


\textbf{Crosswalk.} Mapping path: Census 2010 (OCC2010) $\to$ 6-digit SOC $\to$ O*NET-SOC 2019. Coverage: 882 of 894 O*NET occupations matched (98.7\%). When multiple O*NET codes map to one Census code, distances are averaged.


\subsubsection{Distance Matrix Construction}
\label{subsubsec:distance-construction}


\textbf{Semantic distance ($d_{\mathrm{sem}}$).} Primary measure: Wasserstein distance between occupation activity distributions. Comparison baseline: normalized kernel overlap converted to distance.

\textbf{Wasserstein distance construction:}
\begin{enumerate}
    \item Embed 2,087 DWAs using MPNet (all-mpnet-base-v2, 768 dimensions)
    \item Compute cosine distance matrix $D_{\mathrm{act}}$ over activities (ground metric)
    \item Load occupation-activity importance matrix (894 $\times$ 2,087)
    \item Row-normalize to probability measures: $\rho_j(a) = w_j(a) / \sum_a w_j(a)$
    \item Compute Wasserstein distance for each occupation pair using POT library
\end{enumerate}

\textbf{Kernel overlap construction (comparison baseline):}
\begin{enumerate}
    \item Calibrate kernel bandwidth: $\sigma$ = median nearest-neighbor distance = 0.0096
    \item Build kernel matrix: $K_{ab} = \exp(-D_{\mathrm{act}}(a,b)/\sigma)$
    \item Compute normalized overlap: $O_{ij} = \frac{\sum_{a,b} \rho_i(a) K_{ab} \rho_j(b)}{\sqrt{\sum_{a,b} \rho_i(a) K_{ab} \rho_i(b)} \cdot \sqrt{\sum_{a,b} \rho_j(a) K_{ab} \rho_j(b)}}$
    \item Convert to distance: $d_{\mathrm{sem}}^{\mathrm{kernel}}(i,j) = 1 - O_{ij}$
\end{enumerate}


\textbf{Institutional distance ($d_{\mathrm{inst}}$).} Constructed as job zone difference plus certification barrier:
\[
d_{\mathrm{inst}}(i,j) = |\mathrm{Zone}_i - \mathrm{Zone}_j| + \gamma \cdot |\mathrm{Cert}_i^{\mathrm{norm}} - \mathrm{Cert}_j^{\mathrm{norm}}|
\]

Components:
\begin{itemize}
    \item \textbf{Job Zone} (O*NET ``Job Zones.xlsx''): Scale 1--5 (little preparation $\to$ extensive preparation). Coverage: 923/923 occupations (100\%).
    \item \textbf{Certification importance} (O*NET 30.0 Element ID 2.D.4.a): Scale 1--5 (not important $\to$ extremely important). Coverage: 564/923 occupations (61.1\%). Missing values imputed with median (2.87). Rescaled to 0--4 range.
    \item \textbf{Weight parameter:} $\gamma = 1.0$ (equal weight).
\end{itemize}


\textbf{Correlation between distance measures:}
\begin{itemize}
    \item $\rho(d_{\mathrm{wasserstein}}, d_{\mathrm{inst}}) = 0.249$
    \item $\rho(d_{\mathrm{kernel}}, d_{\mathrm{inst}}) = 0.088$
\end{itemize}
The difference reflects that Wasserstein captures skill transformation costs that correlate with training requirements, while kernel overlap captures shared task mass which is nearly orthogonal to training requirements.


\subsubsection{Conditional Logit Specification}
\label{subsubsec:clogit}


\textbf{Framework:} McFadden's conditional logit for discrete choice.

For worker $k$ who switches from origin occupation $i$, the utility of destination $j$ is:
\[
U_{kj} = \alpha \cdot (-d_{\mathrm{sem}}(i,j)) + \beta \cdot (-d_{\mathrm{inst}}(i,j)) + \varepsilon_{kj}
\]

Distances enter negatively: the model predicts transitions are more likely to destinations with \textbf{lower} semantic and institutional distance from the origin.

\textbf{Choice probability:}
\[
P(j | i, \text{switch}) = \frac{\exp(\alpha \cdot (-d_{\mathrm{sem}}(i,j)) + \beta \cdot (-d_{\mathrm{inst}}(i,j)))}{\sum_{j' \in J} \exp(\alpha \cdot (-d_{\mathrm{sem}}(i,j')) + \beta \cdot (-d_{\mathrm{inst}}(i,j')))}
\]

\textbf{Choice set construction.} Full choice sets (all 447 Census occupations) are computationally prohibitive. Following standard practice, we use random sampling of alternatives: for each observed transition (origin $i \to$ chosen destination $j$), include the chosen destination ($Y = 1$) and sample $K = 10$ non-chosen alternatives uniformly at random ($Y = 0$). This yields 11 observations per transition, grouped by transition ID. Under IIA, random sampling of alternatives yields consistent coefficient estimates.

\textbf{Sample size:}
\begin{itemize}
    \item Observed transitions: 89,329
    \item Rows per transition: 11
    \item Total choice-set observations: 982,619
\end{itemize}


\subsubsection{Results}
\label{subsubsec:cps-results}


\begin{table}[h!]
\centering
\caption{Conditional Logit Results: Wasserstein (Primary)}
\label{tab:clogit-wasserstein}
\begin{tabular}{lcccc}
\toprule
\textbf{Parameter} & \textbf{Coefficient} & \textbf{Std. Error} & \textbf{$t$-statistic} & \textbf{$p$-value} \\
\midrule
$\alpha$ (semantic) & 8.936 & 0.043 & 206.46 & $< 10^{-100}$ \\
$\beta$ (institutional) & 0.142 & 0.004 & 34.59 & $< 10^{-100}$ \\
\bottomrule
\end{tabular}
\end{table}

Log-likelihood: $-183{,}051.4$

Both coefficients are positive (recall covariates are negated distances), indicating transitions are more frequently observed to destinations that are: (1) semantically similar to the origin (lower Wasserstein distance = lower skill transformation cost), and (2) institutionally similar to the origin (similar job zone and certification requirements).


\textbf{Statistical significance.} Both coefficients are estimated with high precision: $\alpha$: $t = 206.5$; $\beta$: $t = 34.6$. The large $t$-statistics primarily reflect sample size (89,329 transitions). Statistical significance at this scale conveys little information about economic importance; we therefore focus interpretation on effect magnitudes and model comparison.


\textbf{Decomposition test.} The key question: Does institutional distance add predictive value beyond semantic distance?

Result: $\beta = 0.142$, $t = 34.59$, $p < 10^{-100}$.

\textbf{Conclusion:} Institutional distance is a significant predictor conditional on semantic distance. Transitions are observed less frequently to destinations with larger job zone gaps and certification barriers, even controlling for Wasserstein-based task transformation cost. The residual institutional component captures non-skill barriers---credentialing gates, licensing, discrimination---that Wasserstein distance does not absorb.


\subsubsection{Interpretation}
\label{subsubsec:cps-interpretation}


\textbf{Consistency with the decomposition.} The results are consistent with the effective distance decomposition (\cref{def:effective-distance}): both $\alpha > 0$ and $\beta > 0$, indicating transitions are less frequently observed to destinations that are either semantically or institutionally distant.


\textbf{Interpretive hypotheses.} If the associations reflect causal mechanisms, the two components would capture distinct barriers:
\begin{enumerate}
    \item \textbf{Semantic distance} ($\alpha = 8.936$): Skill transformation cost---the investment required to reshape one's task portfolio.
    \item \textbf{Institutional distance} ($\beta = 0.142$): Non-skill barriers---credentialing requirements, licensing gates, employer discrimination, network effects.
\end{enumerate}
We cannot distinguish these mechanisms from confounds (e.g., unobserved worker heterogeneity, employer preferences) with the current design.


\textbf{Limitations.}
\begin{enumerate}
    \item \textbf{IIA assumption:} The conditional logit assumes Independence of Irrelevant Alternatives. This assumption is implausible for occupation choice, where adding or removing options should affect relative probabilities of similar occupations. Nested logit or mixed logit specifications would relax this assumption; we retain conditional logit for tractability and comparability with standard practice.
    \item \textbf{Ground metric assumption:} Wasserstein distance inherits assumptions from the ground metric (cosine distance in embedding space). If embeddings misrepresent economic substitutability, Wasserstein distances are correspondingly biased.
    \item \textbf{Panel weights not used:} Estimates are unweighted. Weighted estimation may change point estimates.
    \item \textbf{Choice set sampling:} Random 10:1 sampling of alternatives may introduce bias if the true choice set is systematically different.
    \item \textbf{Certification measure coverage:} Only 61.1\% of occupations have direct certification data; 38.9\% are imputed with the median. This attenuates $\beta$ toward zero; the true institutional effect may be larger.
    \item \textbf{Time period inconsistency:} Distances are computed from 2024 O*NET data but applied to 2015--2024 transitions. Task content and credential requirements may have shifted.
    \item \textbf{Selection on switching:} We observe transitions conditional on switching. Workers who would switch but face prohibitive barriers remain in their origin occupation and are not in the sample.
    \item \textbf{O*NET as constructed perception:} O*NET ratings reflect analyst judgments and incumbent worker surveys, not objective task measurement. The ``task space'' is a constructed representation, not ground truth.
\end{enumerate}


\textbf{Implications for exposure measurement.} If the decomposition is correct, any task-space exposure measure should incorporate both semantic and institutional distance. A model that predicts workers will flow to semantically similar occupations without accounting for institutional barriers would overstate reallocation to credential-protected occupations. This is a conditional claim: it holds if the decomposition reflects real mobility frictions, not merely data artifacts.


\subsubsection{Directional Asymmetry Test}
\label{subsubsec:asymmetric-test}


This section tests whether institutional barriers are asymmetric---whether ``one-way gate'' credentialing theory holds empirically.

\textbf{Hypothesis.} Credentialing theory predicts $\beta_{\mathrm{up}} > \beta_{\mathrm{down}}$, possibly with $\beta_{\mathrm{down}} \approx 0$. Licenses restrict entry but not exit; moving up requires credentials, moving down does not \citep{Jackson2023}.

\textbf{Specification.} Conditional logit with directional institutional distance:
\[
U_{kj} = \alpha \cdot (-d_{\mathrm{sem}}(i,j)) + \beta_{\mathrm{up}} \cdot (-d_{\mathrm{inst}}^{\mathrm{up}}(i,j)) + \beta_{\mathrm{down}} \cdot (-d_{\mathrm{inst}}^{\mathrm{down}}(i,j)) + \varepsilon_{kj}
\]


\begin{table}[h!]
\centering
\caption{Asymmetric Conditional Logit Results (Kernel Specification)}
\label{tab:asymmetric-results}
\begin{tabular}{lcccc}
\toprule
\textbf{Parameter} & \textbf{Coefficient} & \textbf{Std. Error} & \textbf{$t$-statistic} & \textbf{$p$-value} \\
\midrule
$\alpha$ (semantic) & 5.691 & 0.036 & 160.2 & $< 10^{-100}$ \\
$\beta_{\mathrm{up}}$ (upward) & 0.282 & 0.004 & 64.9 & $< 10^{-100}$ \\
$\beta_{\mathrm{down}}$ (downward) & 0.270 & 0.005 & 50.0 & $< 10^{-100}$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Asymmetry ratio:} $\beta_{\mathrm{up}} / \beta_{\mathrm{down}} = 1.04$.

\textbf{Likelihood ratio test:} $\chi^2(1) = 4.33$, $p = 0.038$. The difference is statistically significant but economically trivial.


\textbf{Interpretation: Surprising null.} Contrary to simple credential-gate theory, upward and downward barriers are nearly symmetric. The hypothesis that credentials function as one-way gates is not supported in these data.

\begin{table}[h!]
\centering
\caption{Hypothesis Discrimination: Asymmetric Barriers Test}
\label{tab:asymmetric-hypotheses}
\begin{tabular}{lccc}
\toprule
\textbf{Hypothesis} & \textbf{Prediction} & \textbf{Observed} & \textbf{Consistent?} \\
\midrule
Credentials are one-way gates & $\beta_{\mathrm{up}} \gg \beta_{\mathrm{down}}$ & Ratio $= 1.04$ & No \\
Human capital specificity & $\beta_{\mathrm{up}} \approx \beta_{\mathrm{down}}$ & Ratio $= 1.04$ & Yes \\
Job Zones measure distance, not gates & Symmetric barriers & Symmetric & Yes \\
\bottomrule
\end{tabular}
\end{table}

Three mechanisms explain why barriers appear symmetric:
\begin{enumerate}
    \item \textbf{Overqualification discrimination.} Employers actively penalize overqualified applicants. Audit studies find skills underutilization reduces callbacks with penalties approximately as severe as one year of unemployment \citep{Pedulla2016}. This creates genuine downward barriers.
    
    \item \textbf{Human capital specificity.} Workers forfeit returns to occupation-specific skills when moving down. Estimates suggest 12--20\% wage premiums for five years of occupation tenure \citep{KambourovManovskii2009}, and up to 52\% of overall wage growth derives from task-specific human capital \citep{GathmannSchonberg2010}. Downward moves destroy this accumulated investment.
    
    \item \textbf{Job Zones measure distance, not gates.} O*NET Job Zones capture human capital investment requirements (years of education, training time), not licensing barriers. Licensing creates asymmetric barriers---\citet{Jackson2023} finds licensing reduces entry probability with no significant effect on exit. \citet{KleinerXu2020} find licensing reduces both entry and exit by similar magnitudes. Job Zones capture something more symmetric: general occupational distance.
\end{enumerate}

\textbf{Implication for model space:} Job Zones are valid as distance proxies but do not capture licensing/credential discontinuities. Testing with actual licensing data (e.g., CPS licensing supplement, state-level licensing variation) should reveal the expected gate-like asymmetry \citep{KleinerXu2020}. This is an interpretable null, not a failure---it tells us Job Zones function as a general mobility friction measure, not a credential barrier measure.


\subsection{Complementary Validation: Wage Comovement}
\label{subsec:wage-comovement}


This section reports a complementary validation: whether the semantic geometry predicts wage comovement across occupation pairs. This tests whether the geometry has equilibrium wage implications, not just mobility implications.


\subsubsection{Specification}


\[
\text{WageComovement}_{i,j} = \alpha + \beta \times \text{Overlap}_{i,j} + \varepsilon_{i,j}
\]
Standard errors clustered by origin occupation. $n = 246{,}051$ occupation pairs, $n_{\text{clusters}} = 701$.


\subsubsection{Results}


\begin{table}[h!]
\centering
\caption{Wage Comovement Validation Results}
\label{tab:wage-comovement}
\begin{tabular}{lccccc}
\toprule
\textbf{Measure} & $\beta$ & SE & $t$ & $p$ & $R^2$ \\
\midrule
Normalized kernel overlap & 0.337 & 0.046 & 7.39 & $< 10^{-12}$ & 0.00523 \\
Unnormalized kernel overlap & 1.904 & 0.323 & 5.90 & $< 10^{-8}$ & 0.00310 \\
Binary Jaccard & 0.471 & 0.059 & 8.00 & $< 10^{-15}$ & 0.00167 \\
\bottomrule
\end{tabular}
\end{table}


\subsubsection{Interpretation}


Normalized kernel-weighted overlap achieves $R^2 = 0.00523$, compared to $R^2 = 0.00167$ for binary Jaccard---a threefold relative improvement. However, both values are small in absolute terms: the semantic geometry captures a statistically detectable but quantitatively modest share of wage comovement variance. The value of this test is discriminating between candidate representations, not explaining wage dynamics.

The semantic signal is robust across validation checks:
\begin{itemize}
    \item Semantic overlap is at the 100th percentile of the random baseline distribution (100 seeds).
    \item Permutation $p < 0.001$: the effect is not due to spurious correlation.
    \item After controlling for entropy and support size, the overlap coefficient remains significant ($t = 5.29$).
\end{itemize}

\textbf{Limitations.} Standard errors are clustered by origin occupation. Each occupation appears in many pairs, creating dependence not fully captured by one-way clustering. Two-way clustering (by both origin and destination) would be more conservative and may reduce statistical significance.


\textbf{Relation to CPS mobility findings.} The wage comovement test and the CPS mobility test provide consistent (though not independent) evidence from different data sources:
\begin{itemize}
    \item CPS mobility: worker-level association between distances and destination choice
    \item Wage comovement: occupation-pair correlation in wage changes
\end{itemize}
Both suggest the semantic geometry captures structure correlated with economic outcomes. The CPS test additionally shows that institutional distance is predictive conditional on semantic distance.


\subsection{Application: Automation Prediction}
\label{subsec:automation-prediction}


This section tests whether the semantic geometry improves prediction of employment changes beyond canonical routine task intensity (RTI). This addresses the original RBTC question: can continuous semantic structure add value beyond discrete routine/non-routine classification?


\subsubsection{Motivation}


The framework succeeds at predicting mobility: semantic and institutional distance both predict where workers go when they switch occupations. A natural extension asks whether the same geometry predicts which occupations experience employment declines due to automation. If semantic exposure captures economically relevant structure, it should add predictive value beyond existing measures.


\subsubsection{Data and Measures}


\textbf{Sample:} OES employment data, 2019 and 2024. $N = 658$ occupations after matching to O*NET and external measures. Outcome: $\Delta \log(\text{employment})_{2019 \to 2024}$.

\textbf{Routine Task Intensity (RTI):} Full 16-element Acemoglu-Autor composite \citep{AcemogluAutor2011}, not a single O*NET element. Components include: abstract (quantitative reasoning, creativity, judgment), routine (structured work, repetition), and manual (physical dexterity, outdoor work). RTI = Routine $-$ Abstract $-$ Manual.

\textbf{AI Occupational Exposure (AIOE):} \citet{FeltenRajSeamans2021} measure based on AI capability--ability linkages.

\textbf{Semantic Exposure:} Capability-based shock profile ($v_1$) propagated through task-space kernel geometry.

\textbf{Controls:} Dingel-Neiman telework score \citep{DingelNeiman2020}, O*NET Job Zone.


\subsubsection{RTI Construction Note}


A critical methodological finding: the choice of RTI operationalization matters substantially. Using a single O*NET element (e.g., 4.C.3.b.7 ``Repeating Same Tasks'') yields $R^2 \approx 0$ for employment changes. The full 16-element composite yields $R^2 = 9.8\%$. Proper RTI construction following \citet{AcemogluAutor2011} is essential for meaningful comparison.


\subsubsection{Results}


\begin{table}[h!]
\centering
\caption{Automation Prediction Model Comparison}
\label{tab:automation-results}
\begin{tabular}{lccccc}
\toprule
\textbf{Model} & \textbf{$R^2$} & \textbf{$R^2_{\mathrm{adj}}$} & \textbf{Key $\beta$ (std.)} & \textbf{$p$-value} \\
\midrule
RTI only & 9.82\% & 9.68\% & RTI: $-0.077$ & $< 0.0001$ \\
RTI + Semantic & 12.03\% & 11.76\% & Semantic: $+0.037$ & 0.075 \\
AIOE only & 4.64\% & 4.50\% & AIOE: $+0.053$ & 0.039 \\
AIOE + Semantic & 5.86\% & 5.58\% & Semantic: $+0.033$ & 0.092 \\
Full (RTI + AIOE + Semantic) & 12.23\% & 11.82\% & Only RTI sig. & --- \\
Full + Controls & 12.92\% & 12.25\% & Only RTI sig. & --- \\
\bottomrule
\end{tabular}
\end{table}


\textbf{Incremental validity:} Semantic exposure adds $\Delta R^2 = 2.21\%$ over RTI alone ($p = 0.075$, marginal). In the full model with RTI, AIOE, semantic exposure, and controls, only RTI remains statistically significant.

\textbf{AIOE coefficient sign:} AIOE positively predicts employment changes, consistent with recent evidence on AI complementarity rather than displacement \citep{FeltenRajSeamans2021}.

\textbf{Correlation structure:}
\begin{itemize}
    \item $\mathrm{corr}(\text{RTI}, \text{AIOE}) = -0.325$
    \item $\mathrm{corr}(\text{RTI}, \text{Semantic}) = -0.229$
    \item $\mathrm{corr}(\text{AIOE}, \text{Semantic}) = 0.589$
\end{itemize}
Variance inflation factors are all below 2, indicating no multicollinearity concerns.


\subsubsection{Interpretation}


The framework shows only marginal improvement for predicting automation-driven employment changes. RTI---when properly constructed---dominates. This is an honest partial null that constrains claims about the framework's scope.


\textbf{Productivity-employment gap context.} The \citet{AcemogluRestrepo2018} framework predicts that automation effects on employment are theoretically ambiguous: displacement effects (reducing labor share) compete with reinstatement effects (new tasks increasing demand). RCT evidence shows 14--56\% productivity gains from AI tools \citep{BrynjolfssonLiRaymond2023,NoyZhang2023}; yet administrative data shows precise employment zeros, ruling out effects larger than 1\% \citep{HumlumVestergaard2025}. This disconnect is not anomalous---it reflects an adjustment hierarchy:
\[
\text{Task composition} \to \text{Wages} \to \text{Hours} \to \text{Employment}
\]
Employment responds last. Technology shocks first restructure tasks within occupations, then affect wages, then hours, then headcount. Our marginal result ($\Delta R^2 = 2.2\%$, $p = 0.075$) is consistent with ``too early to measure in employment,'' not ``framework failure.''


\textbf{Substantive interpretation.} Semantic geometry captures task similarity well (validated by mobility and wage comovement), but task similarity is not the same as automation exposure. RTI directly measures the routine/non-routine dimension that theory identifies as the locus of automation effects. Semantic distance captures a broader notion of task relatedness that is useful for mobility prediction but less targeted for automation.


The positive AIOE coefficient suggests current AI technologies may be more complementary than substitutive at the occupation level, consistent with \citet{FeltenRajSeamans2021} findings of positive wage effects for AI-exposed occupations. This is a period-specific finding that may change as AI capabilities evolve.


\textbf{Implication for model space.} Employment may be the wrong outcome variable for near-term technology exposure validation. Task composition changes (e.g., skill mention changes in job postings per \citet{HampoleEtAl2025}) would provide earlier signal. This requires Lightcast/Revelio data---deferred to future work.


\subsection{Retrospective Validation Battery (Pending)}
\label{subsec:retrospective-roadmap}


\begin{center}
\fbox{\parbox{0.9\textwidth}{\textbf{WARNING:} This section specifies tests \emph{not yet executed}. No results are reported. Implementation status: crosswalks specified but not validated.}}
\end{center}


The original research design included three retrospective tests against canonical RBTC settings:
\begin{enumerate}
    \item \textbf{Test A:} Task composition shifts following computerization (ALM setting, 1980--2000)
    \item \textbf{Test B:} Employment reallocation across commuting zones (Autor-Dorn setting, 1980--2005)
    \item \textbf{Test C:} Robot displacement effects (Acemoglu-Restrepo setting, 1990--2007)
\end{enumerate}

These tests were designed before the CPS mobility analysis. We report CPS mobility as the primary test because it operates at the worker level, avoiding ecological inference, and directly tests the semantic-institutional decomposition. The retrospective battery addresses aggregate outcomes (employment shares, wage polarization) that are informative but confounded by industry concentration, geographic clustering, and institutional variation across regions.

\textbf{Status:} Retrospective battery implementation is incomplete. The crosswalks required to map historical Census occupation codes to O*NET-based semantic distances have been specified but not validated. No results from Tests A, B, or C are reported. Full retrospective results will be reported once crosswalk validation is complete.

Full technical specifications, crosswalk documentation, and placeholder implementations are provided in \cref{app:retrospective}.


\subsection{Prospective Evaluation: Generative AI}
\label{subsec:prospective}


The prospective evaluation applies the framework to generative AI. The object is comparison among pre-specified shock profiles, not identification of a single ``true'' profile.


\subsubsection{Candidate Shock Profiles}


Each candidate defines $I^{(v)}(a)$ for activities $a \in T_n$.


\medskip
\noindent\textbf{$v_1$ (capability-only):}
\begin{itemize}
    \item Construction: Score each activity on ``language/cognitive capability relevance'' using a rule-based mapping.
    \item Positive loading on activities tagged as involving information processing, written comprehension, oral expression; negative loading on physical/manual activities.
\end{itemize}


\noindent\textbf{$v_2$ (capability $+$ structure):}
\begin{itemize}
    \item Construction: $v_1$ loadings plus additional positive weight on activities tagged as structured/repetitive.
    \item ``Structured'' operationalized via O*NET ``Structured vs Unstructured'' rating.
    \item Hypothesis: structure amplifies AI exposure beyond raw capability.
\end{itemize}


\noindent\textbf{$v_3$ (external anchor---PRIMARY):}
\begin{itemize}
    \item Construction: Derive from an independent source not using O*NET geometry.
    \item Options: (a) Human expert ratings; (b) Published rubric mapping (Eloundou et al.\ 2024); (c) Patent-activity matching.
    \item Designated as primary profile ex ante.
\end{itemize}


\subsubsection{Profile Selection Protocol}


\noindent\textit{Primary-profile approach (baseline):}
\begin{itemize}
    \item Designate $v_3$ as primary profile ex ante.
    \item $v_1$ and $v_2$ are robustness checks and hypothesis tests.
    \item No profile selection based on outcome fit.
\end{itemize}


\subsubsection{Implications for AI-Based Exposure Measures}


The CPS mobility findings have direct implications for how AI shock profiles should be deployed:

\begin{enumerate}
    \item \textbf{Any task-space oracle must incorporate institutional distance.} Semantic similarity alone overstates the feasibility of worker reallocation. A model that predicts workers displaced by AI will flow to semantically similar occupations without accounting for credentialing barriers will systematically mispredict adjustment paths.
    
    \item \textbf{AI exposure profiles should distinguish ``can do'' from ``allowed to do.''} If generative AI displaces tasks in high-credential occupations (e.g., legal research, medical documentation), workers may face institutional barriers to moving into other occupations even if their semantic skills transfer.
    
    \item \textbf{Counterfactual reallocation simulations should respect $d_{\mathrm{eff}}$.} When simulating how workers will adjust to AI shocks, the oracle should weight destination occupations by $\exp(-d_{\mathrm{eff}})$, not just $\exp(-d_{\mathrm{sem}})$.
\end{enumerate}


\subsection{Empirical Paths Forward}
\label{subsec:paths-forward}


This section specifies pending empirical tests with pre-committed interpretations. Each path follows the template: Test $\to$ Expected Results $\to$ Implications.


\subsubsection{Path A: Optimal Transport Validation --- RESOLVED}


\textbf{Objective:} Test whether Wasserstein distance outperforms kernel overlap for mobility prediction.

\textbf{Status: RESOLVED.} Wasserstein substantially outperforms kernel overlap ($\Delta \text{log-likelihood} = +9{,}576$).

\textbf{Pre-committed interpretation applied:}
\begin{table}[h!]
\centering
\begin{tabular}{lll}
\toprule
\textbf{Outcome} & \textbf{Pre-committed Interpretation} & \textbf{Actual Result} \\
\midrule
OT $\gg$ Kernel & Metric structure matters; adopt OT as primary & \textbf{Confirmed} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Threshold exceeded:} $\Delta \text{log-likelihood} = +9{,}576 \gg 100$ (threshold). $\Delta \alpha = +57\% \gg 10\%$ (threshold).

\textbf{Implication:} Wasserstein is adopted as primary semantic distance measure. The ``earth mover's'' interpretation of occupational distance is economically meaningful: workers minimize skill transformation cost when switching occupations.


\subsubsection{Path B: Embedding Comparison}


\textbf{Objective:} Test whether domain-specific embeddings (JobBERT-v2, trained on ESCO skills) outperform general-purpose embeddings (MPNet).

\textbf{Specification:} Re-compute semantic distances with JobBERT-v2. Compare wage comovement $R^2$ and CPS mobility coefficients.

\textbf{Priority adjustment:} With Wasserstein validated as primary geometry, embedding comparison is lower priority. The geometry change (kernel $\to$ Wasserstein) produced $\Delta \text{LL} = +9{,}576$. The embedding change is unlikely to produce comparable improvement.

\textbf{Pre-committed interpretations:}
\begin{table}[h!]
\centering
\begin{tabular}{lll}
\toprule
\textbf{Outcome} & \textbf{Interpretation} & \textbf{Next Step} \\
\midrule
JobBERT $\gg$ MPNet & Domain-specific training captures economic structure & Adopt JobBERT \\
MPNet $\gg$ JobBERT & General semantic structure sufficient & Stay with MPNet \\
JobBERT $\approx$ MPNet & Embedding choice doesn't matter for this task & Prefer MPNet (standard) \\
\bottomrule
\end{tabular}
\end{table}


\subsubsection{Path C: Retrospective Construct Validity}


\textbf{Objective:} Test whether our semantic distances correlate with Autor-Dorn RTI at occ1990dd level.

\textbf{Specification:} Construct crosswalk from O*NET-SOC to occ1990dd. Compute correlation between semantic exposure and pre-computed RTI from Dorn replication files.

\textbf{Note:} This does NOT require full Autor-Dorn replication---only crosswalk construction and correlation.

\textbf{Pre-committed interpretations:}
\begin{table}[h!]
\centering
\begin{tabular}{lll}
\toprule
\textbf{Outcome} & \textbf{Interpretation} & \textbf{Next Step} \\
\midrule
High correlation ($r > 0.6$) & Semantic geometry captures RTI construct & Validation, not improvement \\
Low correlation ($r < 0.3$) & Semantic geometry captures different structure & Potential contribution \\
Medium correlation & Partial overlap & Need outcome comparison \\
\bottomrule
\end{tabular}
\end{table}


\subsubsection{Path D: Modality-Specific Shocks (Future Work)}


\textbf{Objective:} Construct capability profiles distinguishing code generation, reasoning, agentic workflows. Map to O*NET tasks.

\textbf{Status:} Comprehensive implementation requires 6--9 months (taxonomy design, task annotation, benchmark mapping). One-month pilot (2--3 occupations, $\sim$200 tasks) could establish methodology feasibility.

\textbf{Deferred:} Not feasible in current sprint. Documented as research program extension.


\subsubsection{Path E: Task Composition Outcomes (Future Work)}


\textbf{Objective:} Replace employment with skill mention changes in job postings \citep{HampoleEtAl2025}.

\textbf{Data requirement:} Lightcast/Revelio access (institutional license required).

\textbf{Deferred:} Data not currently available. Documented as research program extension.


\subsubsection{Path F: Asymmetric Test with Wasserstein (Extension)}


\textbf{Objective:} Re-run asymmetric barriers test using Wasserstein distance instead of kernel overlap.

\textbf{Motivation:} The symmetric finding ($\beta_{\mathrm{up}}/\beta_{\mathrm{down}} = 1.04$) was obtained with kernel-based semantic distance. With Wasserstein absorbing more variance from the institutional channel, the residual asymmetry pattern may change.

\textbf{Status:} Natural extension, not yet executed.


\subsection{Model Evaluation Criteria}
\label{subsec:evaluation}


The framework is evaluated against simple baselines using explicit criteria.


\medskip
\noindent\textbf{PRIMARY CRITERIA (all must be addressed):}


\begin{enumerate}
    \item \textbf{Incremental explanatory power.} Activity-based exposure should improve fit relative to discrete baselines by an economically meaningful margin.
    
    \item \textbf{Parameter stability.} Results should be qualitatively stable across domain choices (DWA vs GWA) and weighting schemes.
    
    \item \textbf{Non-redundancy.} We target low-to-moderate correlation between semantic and institutional distance. With Wasserstein, $\rho(d_{\mathrm{sem}}, d_{\mathrm{inst}}) = 0.25$---moderate correlation indicating the measures capture related but distinct structure.
    
    \item \textbf{Non-tautological validation.} Geometry diagnostics must use targets not used in construction.
\end{enumerate}


\medskip
\noindent\textbf{REVISION PROTOCOL:}


If primary criteria fail:
\begin{itemize}
    \item Do not add auxiliary structure to rescue fit.
    \item Report null result: ``activity-based measure does not outperform discrete baselines under tested specifications.''
\end{itemize}


\section{Extensions, Limitations, and Historical Regimes}
\label{sec:extensions}


This paper targets a compact, implementable measurement pipeline. Extensions are organized by status:

\textbf{Completed:}
\begin{itemize}
    \item Wasserstein vs.\ kernel overlap comparison (\cref{subsec:geometry-comparison}): Finding---Wasserstein substantially outperforms ($\Delta \text{LL} = +9{,}576$)
    \item Asymmetric institutional distance (\cref{subsec:asymmetric}): Finding---symmetric barriers
\end{itemize}

\textbf{In Progress:}
\begin{itemize}
    \item Embedding robustness (JobBERT vs.\ MPNet)
    \item Asymmetric test with Wasserstein distance
\end{itemize}

\textbf{Deferred:}
\begin{itemize}
    \item Full retrospective battery (Autor-Dorn replication)
    \item Modality-specific shock profiles
    \item Task composition outcome validation
    \item Structural estimation (Traiberman-style)
\end{itemize}


\subsection{Dynamic accumulation and reversals}
\label{subsec:dynamic}


The baseline construction is static. For settings where shock accumulation or reversal matters, define:
\[
A_t = \delta A_{t-1} + I_t + \varepsilon_t,
\]
where $\delta \in [0,1]$ governs persistence.


\subsection{Asymmetric institutional distance (completed)}
\label{subsec:asymmetric}


The asymmetric extension has been tested (\cref{subsubsec:asymmetric-test}). Contrary to credential-gate theory predicting $\beta^{\mathrm{up}} \gg \beta^{\mathrm{down}}$, we find nearly symmetric barriers: $\beta_{\mathrm{up}} = 0.282$, $\beta_{\mathrm{down}} = 0.270$, ratio $= 1.04$.

This finding discriminates between theoretical models. Human capital specificity theory correctly predicts symmetric barriers: workers accumulate occupation-specific capital lost in any move. Three mechanisms explain why barriers appear symmetric rather than asymmetric:
\begin{enumerate}
    \item \textbf{Occupation-specific human capital} \citep{KambourovManovskii2009}: 12--20\% wage premiums forfeited in any switch
    \item \textbf{Overqualification stigma} \citep{Pedulla2016}: Employers penalize overqualified applicants
    \item \textbf{Job Zones measure preparation, not permission} \citep{KleinerXu2020}: Licensing affects mobility symmetrically
\end{enumerate}

The finding constrains future oracle specifications: effective distance using Job Zones should not require directional asymmetry. However, alternative institutional measures (CPS licensing supplement, state-level licensing variation) may reveal the gate-like asymmetry that simple credentialing theory predicts.


\subsection{Licensing as alternative barrier measure}
\label{subsec:licensing-extension}


Job Zones and occupational licensing measure fundamentally different constructs. Job Zones capture human capital investment requirements---years of education, training time, accumulated experience. Licensing creates legal permission requirements---government authorization to work in an occupation.

\citet{Jackson2023} demonstrates asymmetry using CPS licensing data: licensing reduces entry probability with no significant effect on exit probability. This is the one-way gate pattern that Job Zones fail to capture. \citet{KleinerXu2020} find different results with symmetric effects on both entry and exit---the literature remains unsettled on whether licensing creates one-way or two-way barriers.

Future work should test the decomposition using: (i) CPS licensing supplement (2015+) matched to Institute for Justice policy data; (ii) state-level licensing variation following \citet{KleinerXu2020}; (iii) specific licensed occupations with binding credential requirements (nursing compact variation, bar exam pass rates, medical licensing pathways). These specifications should reveal the asymmetry that simple credentialing theory predicts.


\subsection{Augmentation channel}
\label{subsec:augmentation}


The framework carries $C_t$ (complementarity field) and $E^C$ (augmentation exposure) for symmetry. Estimation requires task-level productivity complementarities not available in O*NET.


\subsection{Supervised dimension extraction}
\label{subsec:supervised}


An alternative to unsupervised text embeddings is supervised extraction of economically-relevant dimensions. Given theoretical priors about dimensions that matter (routine/non-routine, cognitive/manual), one can train linear probes to identify concept directions and project all activities onto these dimensions.


\subsection{Worker mobility as extended validation}
\label{subsec:mobility-extension}


The CPS mobility test uses destination \emph{choice}. An extension would model transition \emph{rates}: do occupation pairs with lower $d_{\mathrm{eff}}$ have higher observed transition probabilities? This tests whether the geometry predicts the \emph{likelihood} of switching, not just the direction conditional on switching.


\subsection{Historical regime signatures}


To summarize historical episodes without forcing a single narrative, \cref{tab:regimes} records regime signatures as hypotheses about shock support, displacement vs augmentation, and primary mechanism.


\begin{table}[h!]
\centering
\begin{tabular}{|l|l|l|l|l|}
\hline
\textbf{Regime} & \textbf{Shock support ($I_t$)} & \textbf{$A$ vs.\ $C$} & \textbf{Mechanism} & \textbf{Wage wedge} \\
\hline
Electrification (1900--1940) & Manual/craft tasks & Mixed & TBD & Context-dependent \\
\hline
Postwar scaling (1945--1970) & Broad/diffuse & Augmentation & TBD & High pass-through \\
\hline
ICT/RBTC (1980--2007) & Routine-cognitive core & Displacement & Both informative & Declining pass-through \\
\hline
AI era (2020--) & Analytic/generative & TBD & TBD & TBD \\
\hline
\end{tabular}
\caption{Historical regimes as parameter signatures.}
\label{tab:regimes}
\end{table}


\section{Conclusion}
\label{sec:conclusion}


This paper develops a geometric framework for measuring occupational mobility and tests it against observed labor market patterns.


\textbf{What we validated.} Task space has metric structure that predicts worker mobility. Wasserstein distance---measuring the minimum cost to transform one occupation's skill portfolio into another---substantially outperforms kernel overlap ($\Delta \text{log-likelihood} = +9{,}576$). This validates the ``earth mover's'' interpretation of occupational distance: workers minimize skill transformation cost when switching occupations.

The semantic-institutional decomposition predicts worker mobility: in 89,329 CPS occupation transitions, both Wasserstein-based semantic distance ($\alpha = 8.936$, $t = 206.5$) and institutional distance ($\beta = 0.142$, $t = 34.6$) are significant predictors of destination choice. The Wasserstein measure correlates moderately with institutional distance ($r = 0.25$), yielding a cleaner decomposition where the residual institutional component captures non-skill barriers---credentialing gates, licensing, discrimination. The semantic geometry predicts wage comovement ($R^2 = 0.52\%$, threefold improvement over binary Jaccard).


\textbf{Informative constraints from nulls.} The asymmetric barriers test yields an interpretable null: $\beta_{\mathrm{up}}/\beta_{\mathrm{down}} = 1.04$. This is an informative constraint on model space, not a negative finding. The symmetric barrier finding rules out pure ``one-way gate'' theories of credentialing: Job Zones capture general occupational distance rather than regulatory gatekeeping. Human capital specificity theory (symmetric barriers from occupation-specific capital) is supported over credential-gate theory (one-way entry barriers).

The automation prediction test shows marginal improvement ($\Delta R^2 = 2.2\%$ over RTI, $p = 0.075$). This too constrains model space: semantic geometry alone is insufficient for near-term automation prediction. The result is consistent with the productivity-employment gap \citep{HumlumVestergaard2025}: employment responds last in the adjustment hierarchy (tasks $\to$ wages $\to$ hours $\to$ employment), so near-term effects appear in task composition before headcount.


\textbf{Model space status.} The research program explores three axes:
\begin{itemize}
    \item \textbf{Representation:} MPNet validated; JobBERT comparison pending
    \item \textbf{Geometry:} Wasserstein validated as primary (substantially outperforms kernel overlap)
    \item \textbf{Shocks:} Generic capability marginal; modality-specific profiles deferred
    \item \textbf{Outcomes:} Mobility validated; employment marginal; task composition deferred
\end{itemize}


\textbf{Theoretical contribution.} The decomposition is grounded in \citet{Traiberman2019}'s switching cost framework, which parameterizes costs as linear in task-space distance with dimension-specific prices. Optimal transport for occupational similarity is validated: Wasserstein distance---the minimum cost to transform one occupation's skill portfolio into another---is economically meaningful. The research program structure is designed for rapid iteration: pre-committed interpretations enable mechanical next-step identification.


These are stress tests of candidate oracle modules. The validated combinations advance; the rejected combinations inform theory. The Wasserstein validation, symmetric barriers finding, and marginal automation result are features, not failures: they identify where candidate representations succeed (mobility prediction with metric structure) and where they fall short (employment prediction in the near term).


\appendix


\section{Retrospective Battery: Technical Specifications}
\label{app:retrospective}


This appendix provides full technical specifications for the retrospective diagnostic battery (Tests A, B, C), which are retained as robustness modules.


\subsection{Test A: Task Composition Shifts (ALM-Informed)}


\textbf{Period:} 1980--2000 (or 1980--1998 to match \citet{AutorLevyMurnane2003} exactly).

\textbf{Core question:} When computerization enters industries, do task composition shifts follow continuous distance from automated tasks, or only binary routine/non-routine classification?

\textbf{Design:}
\begin{itemize}
    \item Unit of analysis: Industry $\times$ decade
    \item Treatment: Computer capital intensity or computer use (CPS October supplements)
    \item Outcome: Change in employment-weighted task intensity by industry
\end{itemize}

\textbf{Key data requirements:}
\begin{itemize}
    \item ALM task measures (DOT 1977 based) or reconstruction via O*NET mapping
    \item Industry-level computer capital/use: BEA capital data, CPS computer use supplements
    \item Industry $\times$ occupation employment matrices: Census 1980, 1990, 2000
    \item Crosswalk: DOT 1977 $\to$ Census occupation $\to$ occ1990dd $\to$ O*NET-SOC $\to$ DWA embeddings
\end{itemize}


\subsection{Test B: Employment Reallocation (Autor-Dorn Setting)}


\textbf{Period:} 1980--2005 (matches \citet{AutorDorn2013} exactly).

\textbf{Core question:} Do commuting zones with higher continuous routine exposure experience different employment polarization patterns than those with equivalent discrete routine share?

\textbf{Design:}
\begin{itemize}
    \item Unit of analysis: Commuting zone ($n = 722$)
    \item Treatment (1980 baseline): Discrete RSH vs.\ continuous CSH
    \item Outcomes: $\Delta$ employment share in service occupations, $\Delta$ employment share in routine occupations, wage changes at percentiles
\end{itemize}

\textbf{Decomposition test:}
\begin{equation}
\Delta Y_{jst} = \delta_t + \beta_1 \times \text{RSH}_{j,t_0} + \beta_3 \times (\text{CSH} - \text{RSH})_{j,t_0} + X'_{j,t_0} \gamma + \phi_s + \varepsilon_{jst}
\end{equation}

If $\beta_3 > 0$ and significant, continuous captures information discrete misses.

\textbf{Replication data sources:}
\begin{itemize}
    \item Autor-Dorn replication archive: \texttt{ddorn.net/data.htm}, File [P2]
    \item Crosswalk files [A1]--[A8]: Census occupation $\to$ occ1990dd (330 balanced codes)
    \item Geographic crosswalks [E1]--[E7]: PUMA/County Group $\to$ CZ with \texttt{afactor} weights
\end{itemize}


\subsection{Test C: Robot Displacement (Acemoglu-Restrepo Setting)}


\textbf{Period:} 1990--2007 (or 1993--2014 depending on IFR data availability).

\textbf{Core question:} Does continuous task overlap with robot-exposed activities predict employment/wage changes better than binary robot exposure classification?

\textbf{Design:}
\begin{itemize}
    \item Unit of analysis: Commuting zone or occupation
    \item Treatment: Industry-level robot exposure (A-R) vs.\ task-level robot exposure (ours)
    \item Outcomes: Employment-to-population ratio, log wages
\end{itemize}

\textbf{Data sources:}
\begin{itemize}
    \item IFR robot data by industry $\times$ country (1993--2014)
    \item A-R replication files
    \item European robot adoption for IV
\end{itemize}


\subsection{Interpretation Matrix}


\begin{table}[h!]
\centering
\begin{tabular}{|l|c|c|c|l|}
\hline
\textbf{Pattern} & \textbf{Test A} & \textbf{Test B} & \textbf{Test C} & \textbf{Implication} \\
\hline
Continuous dominates & $+$ & $+$ & $+$ & H1: Manifold approach preferred throughout \\
\hline
Discrete dominates & $-$ & $-$ & $-$ & H2: Discrete classification sufficient \\
\hline
Adjustment only & $?$ & $+$ & $-$ & H3: Continuous for reallocation, discrete for displacement \\
\hline
Noise & $0$ & $0$ & $0$ & Semantic similarity $\neq$ economic substitutability \\
\hline
\end{tabular}
\caption{Interpretation matrix for retrospective diagnostic battery. ``$+$'' = continuous outperforms discrete; ``$-$'' = discrete outperforms; ``$0$'' = no significant difference.}
\end{table}


% Internal notes removed for external review.
% Distillation planning and venue targeting deferred to internal documentation.


\bibliographystyle{apalike}
\bibliography{references}


\end{document}